Deep Reinforcement Learning for Conservation Decisions

### 2022-01-21 manuscript/manuscript.Rmd
Revise intro to address question of realistic models

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Melissa S. Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mchapman@berkeley.edu
  - name: Kari E. A. Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: kari.norman@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  Can machine learning help us make better decisions about a changing planet?
  In this paper, we illustrate and discuss the potential of a promising 
  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
  most challenging conservation decision problems. RL is uniquely
  well suited to conservation and global change challenges for three reasons: (1)
  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
  utilize rather than replace existing models, simulations, and the knowledge they contain.
  We provide a conceptual and technical introduction to RL and its relevance to ecological
  and conservation challenges, including examples of a problem in setting
  fisheries quotas and in managing ecological tipping points.
  Four appendices with annotated code provide a tangible introduction to researchers 
  looking to adopt, evaluate, or extend these approaches.

keywords:
  - Reinforcement Learning
  - Conservation
  - Machine Learning
  - Artificial Intelligence
  - Tipping points
bibliography: references.bib
nocite: |
    @A2C, @trpo, @ddpg, @TD3, @impala, @hafner_learning_2019, @weber_imagination-augmented_2018
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \usepackage{setspace}
#  - \usepackage{endfloat}
  - \linenumbers
  - \doublespacing

@@

In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.

@@

We illustrate the deep RL approach using the model-free algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
A step-by-step walk-through for training agents on this environment is provided in the Appendix.
We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig. 2].
Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.

@@

The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig. 2].
Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig. 2].
This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
These features highlight a common challenge in the design and training of RL algorithms.
RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
Training these agents in a variety of alternative contexts can improve their ability to generalize to other scenarios.

How could an RL agent be applied to empirical data?
One solution would be to train an agent on a simulation environment that approximates the fishery of interest then query the policy of the agent to find a quota for the observed stock.
To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ramlegacy, see Appendix D].
Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig. 3].
In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.

@@

Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig. 4].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].

@@

Examining the replicate management trajectories and corresponding rewards [Fig. 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady-state strategy.
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
Recall that the quadratic scaling of cost makes the rapid response of the TD3 agent much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.

@@

<!-- RL for behavioral ecology is not the same as RL for conservation -->
This paper does not set the precedent as the first application of RL to ecology.
There have been a number of studies applying RL to behavioral ecology, typically with multi-agent environments [@wang_reinforcement_2020, @frankenhuis_enriching_2019, @perolat_multi-agent_2017].
Yet, it is important to distinguish the aim of these behavioral studies from the aim of applying RL to conservation management. <!-- Compare using RL for conserv as prescriptive, behav eco as descriptive. We don't assume RL is the model. -->
In previous behavioral ecology studies, RL algorithms are used as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
When applying deep RL to conservation management, we do not make the assumption that an RL algorithm learns analogously to how an animal learns.
We instead propose that RL be used as a tool to search for solutions to decision-making problems.

@@

# Acknowledgements {-}

All authors were supported by the NSF Faculty Early Career Development Program Award 1942280. We thank Benjamin Blonder, Zachary Sunberg, and Claire Tomlin for their thoughtful comments.

# Conflicts of Interest {-}

The authors declare there are no conflicts of interest.

# Authors' Contributions {-}

M.L. and C.B. developed the code and wrote the manuscript with support from K.E.A.N. and M.S.C.

# Data Availability {-}

We have created a github repository <https://github.com/boettiger-lab/rl-intro> that contains the code used to produce the figures herein.

>>>>

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Melissa S. Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mchapman@berkeley.edu
  - name: Kari E. A. Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: kari.norman@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  Can machine learning help us make better decisions about a changing planet?
  In this paper, we illustrate and discuss the potential of a promising corner of machine learning known as _reinforcement learning_ (RL) to help tackle the most challenging conservation decision problems.
  RL is uniquely well suited to conservation and global change challenges for three reasons: 
  (1) RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
  (2) RL approaches do not require massive amounts of data, 
  (3) RL approaches would utilize rather than replace existing models, simulations, and the knowledge they contain.
  While this potential may be compelling, the challenges involved in successfully deploying RL-based management to realistic scenarios are formidable --
  the required expertise and computational cost may place these applications may be beyond the reach of all but international technology firms.
  Such an advent would represent a sea-change not only in what problems we can solve, but what stakeholders are involved in solving them.
  Ecologists must establish a better understanding of how these algorithms work and fail if we are to realize this potential and avoid the pitfalls such a transition would bring.
  To this end, we provide a conceptual and technical introduction to RL and its relevance to ecological and conservation challenges, including examples of a problem in setting fisheries quotas and in managing ecological tipping points.
  Four appendices with annotated code provide a tangible introduction to researchers looking to adopt, evaluate, or extend these approaches.

keywords:
  - Reinforcement Learning
  - Conservation
  - Machine Learning
  - Artificial Intelligence
  - Tipping points
bibliography: references.bib
nocite: |
    @A2C, @trpo, @ddpg, @TD3, @impala, @hafner_learning_2019, @weber_imagination-augmented_2018
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
#  - \usepackage{lineno}
#  - \usepackage{endfloat}
#  - \linenumbers

@@

While deep RL may have the potential to open up such intractable problems, it also risks making those problems tractable only by stakeholders with access to extensive computational resources and expertise.
It is notable that the landmark advances cited above have been solved not by academic teams but by specialized research teams of international technology firms such as Alphabet.
Precise estimates of computational resources used in that research are difficult to establish, but previous estimates benchmarked against commercially avilable cloud computing platforms place the training of a single model at over $30 million [@alphacost],
and many realistic ecological problems will involve greater complexity than board games of two players on a 19 x 19 grid [@alphago].
While the history of improved efficiency in computing technology has shown a remarkable ability to reduce such barriers, it has simultaneously moved the leading edge of those capabilities farther beyond reach of traditional ecological research. 
We believe that ecologists must seek to better understand the design, capabilities and limitations of these algorithms, even though their application will surely require the ambitious collaboration, resources and expertise on par with the scale of the immense environmental and ecological problems we face.

In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
We focus on examining the potential and limitations of deep RL through the lens of simple, classical models.
Over a century of theory and practice in ecology has demonstrated that simple models can provide meaningful insights which improve management outcomes [@Getz2017].
As Richard Levins successfully established in his classic paper on the principles of model building [@Levins1966], model complexity must not be mistaken for model realism.
Levins esposed simple mechanistic models which satisfy the goals of being both _realistic_ and _general_. 
More complex models such as those used in fisheries to guide the management of specific stocks typically sacrifice _generality_ for _precision_, a distinction Bob May characterized as "tactical models" (precise) rather than "strategic models" (general). 
Such simple, realistic and general models are still the bedrock of most theory and practice today, even when implementation turns to tactical models for quantitative guidance (for instance, the notion of maximum sustainable yield, MSY, in fisheries, or $R_0$ in epidemiology, remain important concepts in management even though they do not exist in typical tactical models).
These models provide an ideal first benchmark for evaluating the performance of emerging methods of deep RL for several reasons:
Firstly, for some cases the optimal solution is already known, providing a clear standard-of-comparison to evaluate RL performance. 
Prior work sometimes overlooks this essential step, assuming that whatever behavior an RL agent produces is sufficiently optimal or otherwise an accurate proxy for behavior arising from animal learning or evolution. 
As our evaluations will illustrate, such an assumption can be quickly misleading. 
Second, these models are already widely studied and will be familiar to many readers: @Schaefer1954 is a staple of fisheries management textbooks and practice, with over 2800 citations, while @May1977 has become a canonical model of threshholds and tipping points which still continues to dominate how many ecologists think about these phenomena [@Scheffer2009].  
Many readers can thus benefit from exisiting knowledge and intutition about the behavior and implications of these models in interpreting the performance of deep RL, something that would not be possible with a more complex model. 
It would be unwise to consider applications of RL on more complex bespoke models without first analyzing these classic and well-understood cases.
Third, these models include or can easily be extended to contexts for which the optimal management policy is unknown or inaccessible to classical methods.
Our implementations of these models in two python modules published to `PyPi` code archive include many such variations which represent open problems for RL.
We include extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.

We do recognize the need for more complex, "precise" models to derive quanitatively precise management policy in specific management problems such as fisheries, forestry, or climate change mitigation efforts. 
The examples here should provide ecologists with the greater background to engage in the collaborative development of such methods, even as they also highlight challenges which ecological problems pose to existing techniques.
Even the simple models we consider here contain many sources of complexity -- significant stochasticity, non-linear dynamics, hysteresis -- that are likely commonplace to ecological problems but have not been well represtented in RL research to date. 
Such work is beyond the scope of any one paper: it will require not only more computational resources, but more collaborations between large teams of ecologists and computer scientists.

@@

We illustrate the deep RL approach using the model-free algorithm known as Twin Delayed Deep Deterministic Policy Gradient or more simply, TD3 [@TD3].
A step-by-step walk-through for training agents on this environment is provided in the Appendix.
We compare the resulting management, policy, and reward under the RL agent to that achieved by the optimal management solution [Fig 2].
Despite having no knowledge of the underlying model, the RL agent learns enough to achieve nearly optimal performance.

@@

The cumulative reward (utility) realized across 100 stochastic replicates is indistinguishable from that of the optimal policy [Fig 2].
Nevertheless, comparing the mean state over replicate simulations reveals some differences in the RL strategy, wherein the stock is maintained at a slightly higher-than-optimal biomass.
Because our state space and action space are sufficiently low-dimensional in this example, we are also able to visualize the policy function directly, and compare to the optimal policy [Fig 2].
This confirms that quotas tend to be slightly lower than optimal, most notably at larger stock sizes.
These features highlight a common challenge in the design and training of RL algorithms.
RL cares only about improving the realized cumulative reward, and may sometimes achieve this in unexpected ways.
Because these simulations rarely reach stock sizes at or above carrying capacity, these larger stock sizes show a greater deviation from the optimal policy than we observe at more frequently visited lower stock sizes.
Training these agents in a variety of alternative contexts can improve their ability to generalize to other scenarios.

How could an RL agent be applied to empirical data?
One solution would be to train an agent on a simulation environment that approximates the fishery of interest then query the policy of the agent to find a quota for the observed stock.
To illustrate this, we examine the quota that would be recommended by our newly trained RL agent, above, against historical harvest levels of Argentine hake based on stock assessments from 1986 - 2014 [@ramlegacy, see Appendix D].
Hake stocks showed a marked decline throughout this period, while harvests decreased only in proportion [Fig 3].
In contrast, our RL agent would have recommended significantly lower quotas over most of the same interval, including the closure of the fishery as stocks were sufficiently depleted.
While we have no way of knowing for sure if the RL quotas would have led to recovery, let alone an optimal harvest rate, the contrast between those recommended quotas and the historical catch is notable.

@@

Our second example focuses on a case for which we do not have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 4].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of [@May1977] on which our dynamics are based, to microbial dynamics [@Dai2012], lake ecosystem communities [@Carpenter2011] to planetary ecosystems [@Barnosky2012].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $X(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat X_H$, the corresponding ecosystem benefit $b \hat X_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barnosky2012].

@@

Examining the replicate management trajectories and corresponding rewards [Fig 5] reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping well below the mean steady-state reward initially before exceeding it in the long run.
This initial investment then begins to pay off -- by about the 200th time step the RL agent has surpassed the performance of the steady-state strategy.
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.
Recall that the quadratic scaling of cost makes the rapid response of the TD3 agent much more costly to achieve the same net environmental improvement divided into smaller increments over a longer timeline.
However, our RL agent has discovered that the extra investment for a rapid response is well justified as the risk of crossing a tipping point increases.

@@

<!-- RL for behavioral ecology is not the same as RL for conservation -->
This paper does not set the precedent as the first application of RL to ecology.
There have been a number of studies applying RL to behavioral ecology, typically with multi-agent environments [@wang_reinforcement_2020, @frankenhuis_enriching_2019, @perolat_multi-agent_2017].
Yet, it is important to distinguish the aim of these behavioral studies from the aim of applying RL to conservation management. <!-- Compare using RL for conserv as prescriptive, behav eco as descriptive. We don't assume RL is the model. -->
In previous behavioral ecology studies, RL algorithms as a substitute for animal learning mechanisms [@wang_reinforcement_2020, @perolat_multi-agent_2017].
When applying deep RL to conservation management, we do not make the assumption that an RL algorithm learns analogously to how an animal learns.
We instead propose that RL be used as a tool to search for solutions to decision-making problems.

@@
